---
title: "Reporting your best-specified model(s)"
bibliography: handbook.bib

number-sections: true
number-depth: 2

---


```{r} 
#| echo: false
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


```{r echo = FALSE}
# EXAMPLE 1 ----
rm(list=ls())
library(dplyr)

# Example 1: Resp1 ~ Cat1
 n=100
 ss<-sample(c(1:1000), 1)
 set.seed(679) #582
Cat1<-as.factor(sample(c("G", "K", "R"), size=n, replace=TRUE))

library(dplyr)
Group <- as.factor(sample(c("Site1", "Site2", "Site3", "Site4", "Site5", "Site6"),
                           n, replace=TRUE))
uResp<-(as.numeric(Cat1)*4.4+3.9*as.numeric(Group))#+sample(c(200:900), n, replace = TRUE)
Resp<-rnorm(n, uResp, 4.5)
Group <- recode(Group,
                    Site1 = 'Site3',
                    Site2 = 'Site1',
                    Site3 = 'Site5',
                   Site4 = 'Site4',
                   Site5 = 'Site2',
                   Site6 = 'Site6')
Group <- factor(Group, levels = c("Site1","Site2","Site3","Site4","Site5","Site6"))
Cat1 <- recode(Cat1,
                    K = 'Sp1',
                    R = 'Sp2',
                    G = 'Sp3')
Cat1<-factor(Cat1, levels = c("Sp1", "Sp2", "Sp3"))
myDat1<-data.frame(Cat1=Cat1, Resp1=round(Resp,1))
# #write.csv(myDat, "DatEx1.csv", row.names = FALSE)
# 
# library(ggplot2)
# ggplot()+
#   geom_point(data=myDat,
#              mapping=aes(x=Cat1, y=Resp1))
#
startMod<-glm(formula = Resp1 ~ Cat1 + 1, # hypothesis
              data = myDat1, # data
              family = gaussian(link="identity")) # error distribution assumption
# 
# 
# 
library(DHARMa)
# simulationOutput <- simulateResiduals(fittedModel = startMod, n = 250) # simulate data from our model n times
# #
# plot(simulationOutput, asFactor=TRUE) # compare simulated data to our observations
# #
# plot(simulationOutput,
#      form=myDat1$Cat1,
#      asFactor=TRUE) # compare simulated data to our observations
# # #
# myDat$Resid<-simulationOutput$scaledResiduals
# # # # 
# 
# # #
# ggplot()+
#   geom_violin(data=myDat,
#              mapping = aes(x=Group, y=Resid))
# 
# #
# #
library(MuMIn)
options(na.action = "na.fail") # needed for dredge() function to prevent illegal model comparisons
dredgeOut<-dredge(startMod) # fit and compare a model set representing all possible predictor combinations
#
bestMod<-(eval(attr(dredgeOut, "model.calls")$`2`)) # extract model #8 from dredge table
#
# #
# library(emmeans)
# forComp <- emmeans(bestMod, specs =  ~ Cat1, type = "response")
# forComp
# plot(forComp)
# plot(forComp, comparisons = TRUE)
# plot(pairs(emmeans(bestMod, "Cat1"),
#               adjust="scheffe"))

dredgeOut1<-dredgeOut
bestMod1<-bestMod

# EXAMPLE 2 ----


#rm(list=ls())
n=100
ss<-sample(c(1:1000), 1)
set.seed(114) #114
Cat2<-factor(sample(c("TypeA", "TypeB", "TypeC", "TypeD"), size=n, replace=TRUE))
Cat3<-factor(sample(c("Treat1", "Control"), size=n, replace=TRUE), levels=c("Treat1", "Treat2", "Control"))
uResp<-(as.numeric(Cat2)*40.4-33.3*as.numeric(Cat3)+ 23*as.numeric(Cat3)*as.numeric(Cat2))+sample(c(100:300), n, replace = TRUE)
Resp<-rnorm(n, uResp, 4.5)
Cat3[sample(which(Cat3=="Treat1"), round(length(which(Cat3=="Treat1"))/2), replace=TRUE)]<-"Treat2"
Cat2 <- recode(Cat2,
                    TypeC = 'TypeA',
                    TypeD = 'TypeB',
                    TypeB = 'TypeC',
                   TypeA = 'TypeD')
Cat2 <- factor(Cat2, levels = c("TypeA", "TypeB", "TypeC", "TypeD"))

myDat2<-data.frame(Resp2=Resp, Cat2=Cat2, Cat3=Cat3)
# # #write.csv(myDat2, "DatEx2.csv", row.names = FALSE)
# 
# ggplot()+
#   geom_point(data=myDat2,
#              mapping=aes(x=Cat2, y=Resp2, col=Cat3))
#
startMod<-glm(formula = Resp2 ~ Cat2 + Cat3 + Cat2:Cat3, # hypothesis
              data = myDat2, # data
              family = gaussian(link="identity")) # error distribution assumption



# library(DHARMa)
# simulationOutput <- simulateResiduals(fittedModel = startMod, n = 250) # simulate data from our model n times
# #
# plot(simulationOutput, asFactor=TRUE) # compare simulated data to our observations
# #
# plot(simulationOutput,
#      form=myDat2$Cat2,
#      asFactor=TRUE) # compare simulated data to our observations
# 
# plot(simulationOutput,
#      form=myDat2$Cat3,
#      asFactor=TRUE) # compare simulated data to our observations
# 
# # 
# #
# #
library(MuMIn)
options(na.action = "na.fail") # needed for dredge() function to prevent illegal model comparisons
dredgeOut<-dredge(startMod, extra = "R^2") # fit and compare a model set representing all possible predictor combinations
# dredgeOut
#
bestMod<-(eval(attr(dredgeOut, "model.calls")$`8`)) # extract model #8 from dredge table
#
# #
# library(emmeans)
# forComp <- emmeans(bestMod, specs =  ~ Cat1, type = "response")
# forComp
# plot(forComp)
# plot(forComp, comparisons = TRUE)
# plot(pairs(emmeans(bestMod, "Cat1"),
#               adjust="scheffe"))

dredgeOut2<-dredgeOut
bestMod2<-bestMod

# EXAMPLE 3 ----


#rm(list=ls())
n=100
ss<-sample(c(1:1000), 1)
set.seed(261) #261
Num4<-round(runif(n, 0.3, 20.9),2)
uResp<-245*Num4+ 10
Resp<-rnorm(n, uResp, 850)
myDat3<-data.frame(Resp3=Resp, Num4=Num4)

# # #write.csv(myDat3, "DatEx3.csv", row.names = FALSE)


# ggplot()+
#   geom_point(data=myDat3,
#              mapping=aes(x=Num4, y=Resp3))

startMod<-glm(formula = Resp3 ~ Num4 + 1, # hypothesis
              data = myDat3, # data
              family = gaussian(link="identity")) # error distribution assumption


# # 
# library(DHARMa)
# simulationOutput <- simulateResiduals(fittedModel = startMod, n = 250) # simulate data from our model n times
# #
# plot(simulationOutput, asFactor=TRUE) # compare simulated data to our observations
# #
# plot(simulationOutput,
#      form=myDat3$Num4,
#      asFactor=FALSE) # compare simulated data to our observations
# 
# #
# #
# #
library(MuMIn)
options(na.action = "na.fail") # needed for dredge() function to prevent illegal model comparisons
dredgeOut<-dredge(startMod) # fit and compare a model set representing all possible predictor combinations
#dredgeOut
#
bestMod<-(eval(attr(dredgeOut, "model.calls")$`2`)) # extract model #8 from dredge table
#
# #
# library(emmeans)
# forComp <- emmeans(bestMod, specs =  ~ Cat1, type = "response")
# forComp
# plot(forComp)
# plot(forComp, comparisons = TRUE)
# plot(pairs(emmeans(bestMod, "Cat1"),
#               adjust="scheffe"))

dredgeOut3<-dredgeOut
bestMod3<-bestMod

# EXAMPLE 4 ----


#rm(list=ls())
n=100
ss<-sample(c(1:1000), 1)
set.seed(444) #444
Cat5<-as.factor(sample(c("Wild", "Farm", "Urban"), size=n, replace=TRUE))
Num6<-round(runif(n, 300, 700),2)
uResp<-(as.numeric(Cat5)*0.014-0.02*Num6+ as.numeric(Cat5)*0.014*Num6)+100
Resp<-rnorm(n, uResp, 2.5)
myDat4<-data.frame(Resp4=Resp, Cat5=Cat5, Num6=Num6)

# # #write.csv(myDat3, "DatEx3.csv", row.names = FALSE)


# ggplot()+
#   geom_point(data=myDat3,
#              mapping=aes(x=Cont5, y=Resp3, col=Cat4))

startMod<-glm(formula = Resp4 ~ Cat5 + Num6 + Cat5:Num6 + 1, # hypothesis
              data = myDat4, # data
              family = gaussian(link="identity")) # error distribution assumption


# 
# library(DHARMa)
# simulationOutput <- simulateResiduals(fittedModel = startMod, n = 250) # simulate data from our model n times
# #
# plot(simulationOutput, asFactor=TRUE) # compare simulated data to our observations
# #
# plot(simulationOutput,
#      form=myDat3$Cat4,
#      asFactor=TRUE) # compare simulated data to our observations
# 
# plot(simulationOutput,
#      form=myDat3$Cont5,
#      asFactor=FALSE) # compare simulated data to our observations

#
#
# #
library(MuMIn)
options(na.action = "na.fail") # needed for dredge() function to prevent illegal model comparisons
dredgeOut<-dredge(startMod, extra = "R^2") # fit and compare a model set representing all possible predictor combinations
#dredgeOut
#
bestMod<-(eval(attr(dredgeOut, "model.calls")$`8`)) # extract model #8 from dredge table
#
# #
# library(emmeans)
# forComp <- emmeans(bestMod, specs =  ~ Cat1, type = "response")
# forComp
# plot(forComp)
# plot(forComp, comparisons = TRUE)
# plot(pairs(emmeans(bestMod, "Cat1"),
#               adjust="scheffe"))

dredgeOut4<-dredgeOut
bestMod4<-bestMod

```


# Reporting your best-specified models 

::: {.alert .alert-success}
This section is similar regardless of your model structure (e.g. error distribution assumption).  The examples below all assume a normal error distribution assumption but you can use the process presented here for models of any structure.
:::

Reporting your best-specified model means reporting the terms - the predictors and any interactions - that are in your model.  

It is good practice to present the model along with the results from the model selection.  In this way, you can include multiple best-specified models if there is evidence that more than one might be useful.  Depending on your hypothesis and results, you will want to present all models within ∆AICc < 2 of the best-specified model, or all models with any Akaike weight, or simply all models.

Remember from the [Hypothesis Testing](DSPPH_SM_HypothesisTesting.qmd) chapter that you can also use the output from `dredge()` function to report evidence for how you picked the best-specified model.^[more to come in the section on [communicating](DSPPH_SM_Communicating.qmd)]


Let us take a look at how you do this with our examples:


:::{.callout-tip collapse="true" title="Example 1: Resp1 ~ Cat1 + 1"} 

#### Example 1: Resp1 ~ Cat1 + 1


```{r echo = FALSE}

dredgeOut1

```

For Example 1, you will report that your best-specified model is `Resp1 ~ Cat1 + 1`, i.e. that there is evidence that `Cat1` explains variability in `Resp1`.  This was chosen as the best-specified model as it had the lowest AICc and the next highest rank model had a ∆AICc of `r round(dredgeOut1$delta[2],2)` (i.e. ∆AICc > 2).  The best-specified model had an Akaike weight of `r round(dredgeOut1$weight[1],3)`.

:::

:::{.callout-tip collapse="true" title="Example 2: Resp2 ~ Cat2 + Cat3 + Cat2:Cat3 + 1"} 

#### Example 2: Resp2 ~ Cat2 + Cat3 + Cat2:Cat3 + 1

```{r}
#| echo: false


dredgeOut2

```

For Example 2, you will report that your best-specified model is `Resp2 ~ Cat2 + Cat3 + Cat2:Cat3 + 1`, i.e. that there is evidence that `Cat2` and `Cat3` explain variability in `Resp2`, and that there is an interaction between the effect of `Cat2` and `Cat3` on your response - i.e. the effect of `Cat2` on `Resp2` depends on the value of `Cat3`.  

This was chosen as the best-specified model as it had the lowest AICc and the next highest rank model had a ∆AICc of `r round(dredgeOut2$delta[2],2)` (i.e. ∆AICc > 2).  The best-specified model had an Akaike weight of `r round(dredgeOut2$weight[1],3)`.


:::

:::{.callout-tip collapse="true" title="Example 3: Resp3 ~ Num4 + 1"} 

#### Example 3: Resp3 ~ Num4 + 1

```{r}
#| echo: false


dredgeOut3

```
For Example 3, you will report that your best-specified model is `Resp3 ~ Num4 + 1`, i.e. that there is evidence that `Num4` explains variability in `Resp3`.  

This was chosen as the best-specified model as it had the lowest AICc and the next highest rank model had a ∆AICc of `r round(dredgeOut3$delta[2],2)` (i.e. ∆AICc > 2).  The best-specified model had an Akaike weight of `r round(dredgeOut3$weight[1],3)`.

:::


:::{.callout-tip collapse="true" title="Example 4: Resp4 ~ Cat5 + Num6 + Cat5:Num6 + 1"} 

#### Example 4: Resp4 ~ Cat5 + Num6 + Cat5:Num6 + 1

```{r}
#| echo: false


dredgeOut4

```
For Example 4, you will report that your best-specified model is `Resp4 ~ Cat5 + Num6 + Cat5:Num6 + 1`.  This model says that there is evidence that `Cat5` and `Num6` explain variability in `Resp4`, and that there is an interaction between `Cat5` and `Num6` - i.e. the effect of `Num6` on `Resp4` depends on the value of `Cat5`.  

This was chosen as the best-specified model as it had the lowest AICc and the next highest rank model had a ∆AICc of `r round(dredgeOut4$delta[2],2)` (i.e. ∆AICc > 2).  The best-specified model had an Akaike weight of `r round(dredgeOut4$weight[1],3)`.

:::

# [Back to Reporting main page](DSPPH_SM_Reporting.qmd)