---
title: "Statistical Modelling: Reporting"
bibliography: handbook.bib

number-sections: true
number-depth: 2

---

```{r} 
#| echo: false
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


```{r}
#| echo: false
library(png)
library(grid)
library(formatR)
```

::: {.callout-tip collapse="false"} 
## In this chapter you will learn:

- how to report your hypothesis testing results (including visualizations) to communicate what your model says about your hypothesis.

:::

# Reporting your hypothesis testing results

In an upcoming section, we will discuss [communicating your hypothesis testing (including results)](DSPPH_SM_Communicating.qmd) more formally - i.e. the text and visuals you will use to write your scientific reports and papers.  In the current chapter, we will gather many of the pieces needed for that task.

Remember that testing your hypothesis means looking for evidence of an effect of each predictor on your response.  To report your model, you want to communicate this evidence, including:  

## [Reporting your best-specified model](DSPPH_SM_RepBestSpec.qmd)

You will report your best-specified model by reporting the terms (the predictors and any interactions) that are in your model, comparing that to the terms from your starting model.

## [Reporting the explained deviance](DSPPH_SM_RepExpDev.qmd)

You will report how much variability (or deviance) in your response your best-specified model is able to explain, and the relative importance of each model term (each predictor and interaction) in explaining that variability (deviance).

## [Reporting your modelled effects](DSPPH_SM_RepModEffects.qmd)

You will report your modelled effects - i.e. the effect of each predictor on your response.  Remember that model effects are captured in your coefficients. This section will include: 

* how big the effect of a predictor is on your response, 

* how certain you are that there is an effect of the predictor on your response (is the effect bigger than zero?), and 

* (for categorical predictors), if the effects differ among levels of a categorical predictor. 


# Introducing the examples:

In this chapter, we will report model results for four example best-specified models that showcase a range of different hypotheses.  

If you are re-reading this chapter with your own hypothesis you are testing, look for the example that is most similar to your own hypothesis.  

The examples are:  

Example 1: `Resp1 ~ Cat1 + 1`

Example 2: `Resp2 ~ Cat2 + Cat3 + Cat2:Cat3 + 1`

Example 3: `Resp3 ~ Num4 + 1`

Example 4: `Resp4 ~ Cat5 + Num6 + Cat5:Num6 + 1`

where 

* `Resp#` is your response variable. 
* `Cat#` indicates a categorical predictor
* `Num#` indicates a numeric predictor


Note that each example is testing a different hypothesis by fitting a different model to a different data-set.  The four examples are not related to one another. 

Note also that the examples will all involve a normal error distribution assumption.  We will point out (and give examples) when your process would be different for a different error distribution assumption (e.g. Gamma, poisson, and binomial)

Finally, note that we are not going through all the steps that got us to these example best-specified models, but assume we went through the previous steps ([Responses](DSPPH_SM_Responses.qmd), [Predictors](DSPPH_SM_Predictors.qmd), [Hypothesis](DSPPH_SM_Hypothesis.qmd), [Starting model](DSPPH_SM_StartingModel.qmd), [Model validation](DSPPH_SM_ModelValidation.qmd) and [Hypothesis testing](DSPPH_SM_HypothesisTesting.qmd)) as outlined in this Handbook to choose each best-specified model.

:::{.callout-tip collapse="true" title="Example 1: Resp1 ~ Cat1 + 1"} 

#### Example 1: Resp1 ~ Cat1 + 1


```{r echo = FALSE}

rm(list=ls())
library(dplyr)

# Example 1: Resp1 ~ Cat1
 n=100
 ss<-sample(c(1:1000), 1)
 set.seed(679) #582
Cat1<-as.factor(sample(c("G", "K", "R"), size=n, replace=TRUE))

library(dplyr)
Group <- as.factor(sample(c("Site1", "Site2", "Site3", "Site4", "Site5", "Site6"),
                           n, replace=TRUE))
uResp<-(as.numeric(Cat1)*4.4+3.9*as.numeric(Group))#+sample(c(200:900), n, replace = TRUE)
Resp<-rnorm(n, uResp, 4.5)
Group <- recode(Group,
                    Site1 = 'Site3',
                    Site2 = 'Site1',
                    Site3 = 'Site5',
                   Site4 = 'Site4',
                   Site5 = 'Site2',
                   Site6 = 'Site6')
Group <- factor(Group, levels = c("Site1","Site2","Site3","Site4","Site5","Site6"))
Cat1 <- recode(Cat1,
                    K = 'Sp1',
                    R = 'Sp2',
                    G = 'Sp3')
Cat1<-factor(Cat1, levels = c("Sp1", "Sp2", "Sp3"))
myDat1<-data.frame(Cat1=Cat1, Resp1=round(Resp,1))
# #write.csv(myDat, "DatEx1.csv", row.names = FALSE)
# 
# library(ggplot2)
# ggplot()+
#   geom_point(data=myDat,
#              mapping=aes(x=Cat1, y=Resp1))
#
startMod<-glm(formula = Resp1 ~ Cat1 + 1, # hypothesis
              data = myDat1, # data
              family = gaussian(link="identity")) # error distribution assumption
# 
# 
# 
library(DHARMa)
# simulationOutput <- simulateResiduals(fittedModel = startMod, n = 250) # simulate data from our model n times
# #
# plot(simulationOutput, asFactor=TRUE) # compare simulated data to our observations
# #
# plot(simulationOutput,
#      form=myDat1$Cat1,
#      asFactor=TRUE) # compare simulated data to our observations
# # #
# myDat$Resid<-simulationOutput$scaledResiduals
# # # # 
# 
# # #
# ggplot()+
#   geom_violin(data=myDat,
#              mapping = aes(x=Group, y=Resid))
# 
# #
# #
library(MuMIn)
options(na.action = "na.fail") # needed for dredge() function to prevent illegal model comparisons
dredgeOut<-dredge(startMod) # fit and compare a model set representing all possible predictor combinations
#
bestMod<-(eval(attr(dredgeOut, "model.calls")$`2`)) # extract model #8 from dredge table
#
# #
# library(emmeans)
# forComp <- emmeans(bestMod, specs =  ~ Cat1, type = "response")
# forComp
# plot(forComp)
# plot(forComp, comparisons = TRUE)
# plot(pairs(emmeans(bestMod, "Cat1"),
#               adjust="scheffe"))

dredgeOut1<-dredgeOut
bestMod1<-bestMod

```

For example #1, assume your best-specified model says that your response variable (`Resp1`) is explained by a categorical predictor (`Cat1`):    

`Resp1 ~ Cat1`

Your best-specified model for example #1 is in an object called `bestMod1`:

```{r}

bestMod1

```
that was fit to data in `myDat1`:

```{r}

str(myDat1)

```

and the dredge() table you used to pick your bestMod1 is in `dredgeOut1`:

```{r}

dredgeOut1

```
:::

:::{.callout-tip collapse="true" title="Example 2: Resp2 ~ Cat2 + Cat3 + Cat2:Cat3 + 1"} 


#### Example 2: Resp2 ~ Cat2 + Cat3 + Cat2:Cat3 + 1


```{r echo = FALSE}

#rm(list=ls())
n=100
ss<-sample(c(1:1000), 1)
set.seed(114) #114
Cat2<-factor(sample(c("TypeA", "TypeB", "TypeC", "TypeD"), size=n, replace=TRUE))
Cat3<-factor(sample(c("Treat1", "Control"), size=n, replace=TRUE), levels=c("Treat1", "Treat2", "Control"))
uResp<-(as.numeric(Cat2)*40.4-33.3*as.numeric(Cat3)+ 23*as.numeric(Cat3)*as.numeric(Cat2))+sample(c(100:300), n, replace = TRUE)
Resp<-rnorm(n, uResp, 4.5)
Cat3[sample(which(Cat3=="Treat1"), round(length(which(Cat3=="Treat1"))/2), replace=TRUE)]<-"Treat2"
Cat2 <- recode(Cat2,
                    TypeC = 'TypeA',
                    TypeD = 'TypeB',
                    TypeB = 'TypeC',
                   TypeA = 'TypeD')
Cat2 <- factor(Cat2, levels = c("TypeA", "TypeB", "TypeC", "TypeD"))

myDat2<-data.frame(Resp2=Resp, Cat2=Cat2, Cat3=Cat3)
# # #write.csv(myDat2, "DatEx2.csv", row.names = FALSE)
# 
# ggplot()+
#   geom_point(data=myDat2,
#              mapping=aes(x=Cat2, y=Resp2, col=Cat3))
#
startMod<-glm(formula = Resp2 ~ Cat2 + Cat3 + Cat2:Cat3, # hypothesis
              data = myDat2, # data
              family = gaussian(link="identity")) # error distribution assumption



# library(DHARMa)
# simulationOutput <- simulateResiduals(fittedModel = startMod, n = 250) # simulate data from our model n times
# #
# plot(simulationOutput, asFactor=TRUE) # compare simulated data to our observations
# #
# plot(simulationOutput,
#      form=myDat2$Cat2,
#      asFactor=TRUE) # compare simulated data to our observations
# 
# plot(simulationOutput,
#      form=myDat2$Cat3,
#      asFactor=TRUE) # compare simulated data to our observations
# 
# # 
# #
# #
library(MuMIn)
options(na.action = "na.fail") # needed for dredge() function to prevent illegal model comparisons
dredgeOut<-dredge(startMod, extra = "R^2") # fit and compare a model set representing all possible predictor combinations
# dredgeOut
#
bestMod<-(eval(attr(dredgeOut, "model.calls")$`8`)) # extract model #8 from dredge table
#
# #
# library(emmeans)
# forComp <- emmeans(bestMod, specs =  ~ Cat1, type = "response")
# forComp
# plot(forComp)
# plot(forComp, comparisons = TRUE)
# plot(pairs(emmeans(bestMod, "Cat1"),
#               adjust="scheffe"))

dredgeOut2<-dredgeOut
bestMod2<-bestMod

```

For Example #2, assume your best-specified model says that your response variable (`Resp2`) is explained by two categorical predictors (`Cat2` & `Cat3`) as well as the interaction between the predictors:    

`Resp2 ~ Cat2 + Cat3 + Cat2:Cat3 + 1`

Your best-specified model for example #2 is in an object called `bestMod2`:

```{r}

bestMod2

```

that was fit to data in `myDat2`:

```{r}

str(myDat2)

```

and the `dredge()` table you used to pick your `bestMod2` is in `dredgeOut2`:

```{r}

dredgeOut2

```


:::

:::{.callout-tip collapse="true" title="Example 3: Resp3 ~ Num4 + 1"} 

#### Example 3: Resp3 ~ Num4 + 1


```{r echo = FALSE}

#rm(list=ls())
n=100
ss<-sample(c(1:1000), 1)
set.seed(261) #261
Num4<-round(runif(n, 0.3, 20.9),2)
uResp<-245*Num4+ 10
Resp<-rnorm(n, uResp, 850)
myDat3<-data.frame(Resp3=Resp, Num4=Num4)

# # #write.csv(myDat3, "DatEx3.csv", row.names = FALSE)


# ggplot()+
#   geom_point(data=myDat3,
#              mapping=aes(x=Num4, y=Resp3))

startMod<-glm(formula = Resp3 ~ Num4 + 1, # hypothesis
              data = myDat3, # data
              family = gaussian(link="identity")) # error distribution assumption


# # 
# library(DHARMa)
# simulationOutput <- simulateResiduals(fittedModel = startMod, n = 250) # simulate data from our model n times
# #
# plot(simulationOutput, asFactor=TRUE) # compare simulated data to our observations
# #
# plot(simulationOutput,
#      form=myDat3$Num4,
#      asFactor=FALSE) # compare simulated data to our observations
# 
# #
# #
# #
library(MuMIn)
options(na.action = "na.fail") # needed for dredge() function to prevent illegal model comparisons
dredgeOut<-dredge(startMod) # fit and compare a model set representing all possible predictor combinations
#dredgeOut
#
bestMod<-(eval(attr(dredgeOut, "model.calls")$`2`)) # extract model #8 from dredge table
#
# #
# library(emmeans)
# forComp <- emmeans(bestMod, specs =  ~ Cat1, type = "response")
# forComp
# plot(forComp)
# plot(forComp, comparisons = TRUE)
# plot(pairs(emmeans(bestMod, "Cat1"),
#               adjust="scheffe"))

dredgeOut3<-dredgeOut
bestMod3<-bestMod

```

For Example #3, assume your best-specified model says that your response variable (`Resp3`) is explained by one numeric predictor (`Num4`):    

`Resp3 ~ Num4 + 1`

Your best-specified model for example #3 is in an object called `bestMod4`:

```{r}

bestMod3

```

that was fit to data in `myDat4`:

```{r}

str(myDat3)

```

and the `dredge()` table you used to pick your `bestMod3` is in `dredgeOut3`: 

```{r}

dredgeOut3

```
:::

:::{.callout-tip collapse="true" title="Example 4: Resp4 ~ Cat5 + Num6 + Cat5:Num6 + 1"} 

#### Example 4: Resp4 ~ Cat5 + Num6 + Cat5:Num6 + 1

```{r echo = FALSE}

#rm(list=ls())
n=100
ss<-sample(c(1:1000), 1)
set.seed(444) #444
Cat5<-as.factor(sample(c("Wild", "Farm", "Urban"), size=n, replace=TRUE))
Num6<-round(runif(n, 300, 700),2)
uResp<-(as.numeric(Cat5)*0.014-0.02*Num6+ as.numeric(Cat5)*0.014*Num6)+100
Resp<-rnorm(n, uResp, 2.5)
myDat4<-data.frame(Resp4=Resp, Cat5=Cat5, Num6=Num6)

# # #write.csv(myDat3, "DatEx3.csv", row.names = FALSE)


# ggplot()+
#   geom_point(data=myDat3,
#              mapping=aes(x=Cont5, y=Resp3, col=Cat4))

startMod<-glm(formula = Resp4 ~ Cat5 + Num6 + Cat5:Num6 + 1, # hypothesis
              data = myDat4, # data
              family = gaussian(link="identity")) # error distribution assumption


# 
# library(DHARMa)
# simulationOutput <- simulateResiduals(fittedModel = startMod, n = 250) # simulate data from our model n times
# #
# plot(simulationOutput, asFactor=TRUE) # compare simulated data to our observations
# #
# plot(simulationOutput,
#      form=myDat3$Cat4,
#      asFactor=TRUE) # compare simulated data to our observations
# 
# plot(simulationOutput,
#      form=myDat3$Cont5,
#      asFactor=FALSE) # compare simulated data to our observations

#
#
# #
library(MuMIn)
options(na.action = "na.fail") # needed for dredge() function to prevent illegal model comparisons
dredgeOut<-dredge(startMod, extra = "R^2") # fit and compare a model set representing all possible predictor combinations
#dredgeOut
#
bestMod<-(eval(attr(dredgeOut, "model.calls")$`8`)) # extract model #8 from dredge table
#
# #
# library(emmeans)
# forComp <- emmeans(bestMod, specs =  ~ Cat1, type = "response")
# forComp
# plot(forComp)
# plot(forComp, comparisons = TRUE)
# plot(pairs(emmeans(bestMod, "Cat1"),
#               adjust="scheffe"))

dredgeOut4<-dredgeOut
bestMod4<-bestMod

```

For Example #4, assume your best-specified model says that your response variable (`Resp4`) is explained by one categorical predictor (`Cat4`) and one numeric predictor (`Num6`) as well as the interaction between the predictors:    

`Resp4 ~ Cat5 + Num6 + Cat5:Num6 + 1`

Your best-specified model for example #3 is in an object called `bestMod4`:

```{r}

bestMod4

```

that was fit to data in `myDat4`:

```{r}

str(myDat4)

```

and the `dredge()` table you used to pick your `bestMod4` is in `dredgeOut4`:

```{r}

dredgeOut4

```
:::



# Up next {.unnumbered}

In the next chapter, we will discuss how you can use your model to make [predictions](DSPPH_SM_Predicting.qmd). 



